---
title: "Pratical Machine Learning Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).


### Data

The training data for this project are available here: <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here: <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>


The data for this project come from this source: <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har>.

### What you should submit

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. I may use any of the other variables to predict with. I should create a report describing how I built my model, how I used cross validation, what I think the expected out of sample error is, and why I made the choices I did. I will also use my prediction model to predict 20 different test cases.


## 2. Analysis

### Load the library and seed

```{r, warning=FALSE, message=FALSE}
library(lattice)
library(ggplot2)
library(caret)
library(randomForest) #Random forest for classification and regression
library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot

# setting the overall seed for reproduceability
set.seed(1234)
```

### Load the data and preliminary cleaning

First we want to load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.
Results will be hidden from the report for clarity and space considerations.

```{r, results='hide'}
# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  

# Loading the training data set into my R session replacing all missing with "NA"
filename <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
training <- read.csv(url(filename), na.strings=c("NA","#DIV/0!",""))

# Loading the test data set into my R session replacing all missing with "NA"
filename <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing <- read.csv(url(filename), na.strings=c("NA","#DIV/0!",""))

# Check dimensions for number of variables and number of observations
dim(training) ; dim(testing)

# Delete columns with all missing values
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]
```

```{r}
# and have a look at our new datasets:
dim(training) ; dim(testing)
```
```{r, results='hide'}
head(training) ; head(testing)
```
After the data cleaning the variables selected to specify the model are:
```{r}
names(training)
```

## 3. Model build

### Cross-validation
Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (80% of the original Training data set) and subTesting data (20%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

```{r}
subsamples <- createDataPartition(y=training$classe, p=0.80, list=FALSE)
subTraining <- training[subsamples, ] 
subTesting <- training[-subsamples, ]
dim(subTraining) ; dim(subTesting)
```

### Look at the data

The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r}
plot(subTraining$classe, col="lightblue", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")
```

From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

### Expected out-of-sample error
The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

### Prediction model 1 : Using Random Forest
```{r}
model1 <- randomForest(classe ~. , data=subTraining, method="class")

# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction1, subTesting$classe)
```

### Prediction model 2 : Using Decision Tree
```{r}
model2 <- rpart(classe ~ ., data=subTraining, method="class")

# Predicting:
prediction2 <- predict(model2, subTesting, type = "class")

# Plot of the Decision Tree
rpart.plot(model2, main="Classification Tree", extra=102, under=TRUE, faclen=0)

# Test results on our subTesting data set:
confusionMatrix(prediction2, subTesting$classe)
```


### 4. Conclusion 
Finally, Random Forest algorithm performed better than Decision Trees. Accuracy for Random Forest model was 0.997 compared to 0.751 for Decision Tree model. 

The random Forest model is choosen. The accuracy of the model is 0.997. The expected out-of-sample error is estimated at 0.003, or 0.3%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. 

Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

### 5. Submission
Here is the final outcome based on the Prediction Model 1 (Random Forest) applied against the Testing dataset :
```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model1, testing, type="class")
predictfinal
```

